---
title: "Introduction to amRml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to amRml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(amRml)
```

## Overview

`amRml` is part of a 3-package suite for predicting antimicrobial resistance (AMR) using machine learning models trained on bacterial genomic features.

This vignette demonstrates how to generate ML input matrices, train models, and visualize results.

## Generating ML input matrices

After data curation with `amRdata`, use `generateMLInputs()` to create ML-ready matrices from the DuckDB containing metadata and feature parquets.
The DuckDB with the parquet views is created with `cleanData()`

```{r generate-matrices}
# Example uses Shigella flexneri (Sfl); replace paths with your own species and drug.
generateMLInputs(
  parquet_duckdb_path = "results/Sfl_parquet.duckdb",
  out_path = "results/",
  n_fold = 5,
  split = c(1, 0), # CV mode (or use split = 0 as shorthand)
  min_n = 25,
  verbosity = "minimal"
)
```

This generates:

- Drug and drug-class matrices for all feature types (genes, proteins, domains, struct)
- Year-stratified and country-stratified matrices
- Leave-one-out (LOO) matrices for temporal/geographic holdouts
- Multi-drug resistance (MDR) matrices

For classical train/validation/test splits instead of cross-validation:

```{r generate-splits}
generateMLInputs(
  parquet_duckdb_path = "results/Sfl_parquet.duckdb",
  out_path = "results/",
  n_fold = NULL,
  split = c(0.7, 0.15),
  verbosity = "debug"
)
```

Use `skipImbalancedMatrix()` to check if a drug-bug combination has sufficient samples:

```{r skip-check}
skip <- skipImbalancedMatrix(
  genome_ids = genome_ids,
  phenotype_counts = phenotype_counts,
  n_fold = 5,
  split = c(1, 0),
  min_total_obs = 40
)
```

## Loading ML matrices

The package expects Parquet files in long (sparse) format with these columns:

| Column | Description |
|--------|-------------|
| `genome_id` | Unique identifier for each isolate |
| `feature_id` | Feature name (gene, protein, domain, or struct) |
| `value` | Binary presence/absence (0 or 1) or count |
| `genome_drug.resistant_phenotype` | "Resistant" or "Susceptible" |

`loadMLInputTibble()` converts this to wide format (one row per genome, one column per feature) for ML.

```{r load-data}
ml_tibble <- loadMLInputTibble(
  parquet_path = "results/matrix/Sfl_drug_AMP_domains_binary_sparse.parquet"
)

# Check dimensions
n_features <- getNumFeat(ml_tibble)
target_var <- getTargetVarName(ml_tibble)
```

## Main pipeline

The primary entry point is `runMLPipeline()`, which orchestrates the entire ML workflow:

```{r run-pipeline}
results <- runMLPipeline(
  ml_input_tibble = ml_tibble,
  model = "LR",
  split = c(0.6, 0.2),
  n_fold = 2,
  n_top_feats = 20,
  select_best_metric = "mcc",
  return_fit = TRUE,
  return_pred = TRUE,
  verbose = TRUE
)

# View results
results$performance_tibble
results$top_feat_tibble
```

### Output structure

`runMLPipeline()` returns a list with:

**`performance_tibble`** - model performance metrics:

| Column | Description |
|--------|-------------|
| `num_obs` | Number of observations |
| `res_prop` | Proportion of resistant samples |
| `n_feat` | Number of features |
| `model` | Model type (LR) |
| `train_prop`, `val_prop` | Train/validation split proportions |
| `fit_penalty`, `fit_mixture` | Fitted hyperparameters |
| `mcc`, `nmcc`, `f1`, `bal_acc`, `log2_apop` | Performance metrics |
| `run_time_sec` | Runtime in seconds |

**`top_feat_tibble`** - ranked feature importance:

| Column | Description |
|--------|-------------|
| `Variable` | Feature name |
| `Importance` | Variable importance score |
| `Sign` | Direction of effect (POS = resistance, NEG = susceptibility) |

**Optional outputs** (when `return_*` = TRUE):

- `tune_res` - tuning results from grid search
- `fit` - fitted model workflow object
- `pred` - predictions with `.pred_class`, `.pred_Resistant`, `.pred_Susceptible`

## Step-by-step model building

For more control, use the individual tidymodels-based functions:

```{r split-data}
# Split data
data_split <- splitMLInputTibble(ml_tibble, split = c(0.6, 0.2), seed = 123)

library(rsample)
train_data <- training(data_split)
test_data <- testing(data_split)
```

```{r build-model}
# Build recipe (preprocessing)
recipe <- buildRecipe(train_data, use_pca = FALSE, pca_threshold = 0.95)

# Build logistic regression model
lr_model <- buildLRModel(multi_class = FALSE)

# Build workflow
wflow <- buildWflow(lr_model, recipe)

# Build tuning grid
grid <- buildTuningGrid(
  model = "LR",
  penalty_vec = 10^seq(-4, -1, length.out = 10),
  mix_vec = 0:5 / 5
)
```

```{r tune-fit}
# Tune hyperparameters
tune_res <- tuneGrid(wflow, data_split, grid, n_fold = 5)

# Select and fit best model
best_wflow <- selectBestModel(tune_res, wflow, select_best_metric = "mcc")
fit <- fitBestModel(best_wflow, train_data)

# Get fitted hyperparameters
fit_hps <- getFitHps(fit)

# Make predictions
predictions <- predict(fit, test_data)
```

## Performance metrics

```{r metrics}
# Individual metrics
mcc <- calculateMCC(predictions) # Matthews correlation coefficient (-1 to 1)
nmcc <- calculatenMCC(predictions) # Normalized MCC (0 to 1)
f1 <- calculateF1(predictions) # F1 score
bal_acc <- calculateBalAcc(predictions) # Balanced accuracy
auprc <- calculateAUPRC(predictions) # Area under PR curve
log2_apop <- calculateLog2APOP(predictions) # log2(AUPRC/prior)

# All metrics at once
all_metrics <- calculateEvalMets(predictions)

# Confusion matrix
conf_mat <- getConfusionMatrix(predictions)
```

| Metric | Function | Description |
|--------|----------|-------------|
| MCC | `calculateMCC()` | Matthews correlation coefficient (-1 to 1) |
| nMCC | `calculatenMCC()` | Normalized Matthews correlation coefficient (0 to 1) |
| F1 | `calculateF1()` | Harmonic mean of precision and recall |
| Balanced accuracy | `calculateBalAcc()` | Average of sensitivity and specificity |
| AUPRC | `calculateAUPRC()` | Area under precision-recall curve |
| log2(AUPRC/prior) | `calculateLog2APOP()` | Class-imbalance adjusted AUPRC |

## Feature importance

```{r feature-importance}
# Extract top features from fitted model by number
top_features <- extractTopFeats(fit, n_top_feats = 20)

# Or by proportion of variable importance
top_features <- extractTopFeats(fit, prop_vi_top_feats = c(0, 0.2))

# Columns: Variable, Importance, Sign
# Positive Sign = associated with resistance
# Negative Sign = associated with susceptibility
```

## Iterative feature elimination (IFE)

IFE iteratively removes top features to identify the most predictive subset:

```{r ife}
ife_results <- runIFE(
  ml_tibble,
  by_num = TRUE, # Remove by number of features
  by_vi = FALSE, # Or remove by variable importance
  percent_removal_vec = 10 * 1:9, # Remove 10%, 20%, ..., 90%
  mix_vec = 0, # L2 regularization for IFE
  return_feats = TRUE,
  verbose = TRUE
)

# Results include MCC and nMCC at each iteration
ife_results$ife_performance_tibble
ife_results$feats_removed

# Remove specific features manually
ml_tibble_reduced <- removeTopFeats(ml_tibble, top_features)
```

## Visualization

### Precision-recall curve

```{r plot-prc}
test_data_plus_predictions <- readr::read_tsv("results/ML_pred/Sfl_drug_AMP_domains_binary_prediction.tsv")
plotPRC(test_data_plus_predictions)
```
### ROC curve

```{r plot-roc}
test_data_plus_predictions <- readr::read_tsv("results/ML_pred/Sfl_drug_AMP_domains_binary_prediction.tsv")
plotROC(test_data_plus_predictions)
```
### Variable importance plot

```{r plot-vi}
topfeat <- readr::read_tsv("results/ML_top_features/Sfl_drug_AMP_domains_binary_top_features.tsv")
plotTopFeatsVI(topfeat)
```
### Baseline comparison barplot
non_shuffled_label_results Output of `runMLPipeline(shuffle_labels = FALSE)`
shuffled_label_results Output of `runMLPipeline(shuffle_labels = TRUE)`
```{r plot-baseline}
# Run with shuffled labels for baseline
baseline_results <- runMLPipeline(
  ml_input_tibble = ml_tibble,
  model = "LR",
  split = c(0.6, 0.2),
  shuffle_labels = TRUE
)


# Compare performance
getBaselineComparisonBarplot(
  non_shuffled_label_results = results$performance_tibble,
  shuffled_label_results = baseline_results$performance_tibble
)
```

## Fisher's exact tests (baseline comparison)

As a non-ML baseline, run Fisher's exact tests with multiple testing correction:

```{r fisher}
# Complete Fisher pipeline
fisher_results <- runFishers(
  matrix_path = "results/matrix/Sfl_drug_AMP_genes_binary_sparse.parquet",
  Q = 0.05,
  alternative = "two.sided",
  susceptible_label = "Susceptible",
  resistant_label = "Resistant"
)

# Or step-by-step:
encoded <- encodePhenotype(ml_tibble)
fisher_res <- runFisherTests(encoded$df, encoded$target, alternative = "two.sided")
fisher_res <- applyBenjaminiHochberg(fisher_res, Q = 0.05)
fisher_res <- computeFeatureFreq(encoded$df, fisher_res, encoded$target)
```

### Plot Fisher's results
You can label the top N features to highlight the strongest hits (default is 5) or change the BH-adjusted p-value threshold with the alpha argument

```{r}
plotFishers(fisher_results)
plotFishers(fisher_results, alpha = 0.01, label_top_n = 5)
```

## Wrapper to run all models 
Given a DuckDB file produced by `runDataProcessing()`, it:
1. generates all ML feature matrices (drug, class, year, country, MDR, LOO)
2. creates all ML directory structures
3. prepares ML input lists for every mode
4. runs logistic regression ML models (standard + stratified + MDR)
5. saves performance metrics, fitted models, predictions, and top feature rankings
``` {r}
runModelingPipeline(parquet_duckdb_path,
  threads = 16,
  n_fold = 5,
  split = c(1, 0),
  min_n = 25,
  prop_vi_top_feats = c(0, 1),
  pca_threshold = 0.99,
  verbose = TRUE,
  use_saved_split = TRUE
)
```

Merge the performance and top features of each kind of models into a parquet that will serve as starting data for `amRshiny` package

``` {r}
buildPerformancePq(
  path = "results/ML_perf",
  stratify_by = NULL,
  LOO = FALSE,
  MDR = FALSE,
  cross_test = FALSE,
  out_parquet = NULL,
  compression = "zstd",
  verbose = TRUE
)
buildTopFeatsPq(
  path = "results/ML_top_features",
  stratify_by = NULL,
  LOO = FALSE,
  MDR = FALSE,
  cross_test = FALSE,
  out_parquet = NULL,
  compression = "zstd",
  verbose = TRUE
)
```
