---
title: "Introduction to amRml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to amRml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(amRml)
```

## Overview

`amRml` is part of a 3-package suite for predicting antimicrobial resistance (AMR) using machine learning models trained on bacterial genomic features.

This vignette demonstrates how to generate ML input matrices, train models, and visualize results.

## Generating ML input matrices

After data curation with `amRdata`, use `generateMLInputs()` to create ML-ready matrices from the DuckDB containing metadata and feature parquets.

```{r generate-matrices}
# Generate all ML input matrices from curated data
generateMLInputs(
  parquet_duckdb_path = "results/Cje_parquet.duckdb",
  out_path = "results/",
  n_fold = 5,
  split = c(1, 0), # CV mode (or use split = 0 as shorthand)
  min_n = 25,
  verbosity = "minimal"
)
```

This generates:

- Drug and drug-class matrices for all feature types (genes, proteins, domains, struct)
- Year-stratified and country-stratified matrices
- Leave-one-out (LOO) matrices for temporal/geographic holdouts
- Multi-drug resistance (MDR) matrices

For classical train/validation/test splits instead of cross-validation:

```{r generate-splits}
# 70% train, 15% validation, 15% test
generateMLInputs(
  parquet_duckdb_path = "results/Cje_parquet.duckdb",
  out_path = "results/",
  n_fold = NULL,
  split = c(0.7, 0.15),
  verbosity = "debug"
)
```

Use `skipImbalancedMatrix()` to check if a drug-bug combination has sufficient samples:

```{r skip-check}
skip <- skipImbalancedMatrix(
  genome_ids = genome_ids,
  phenotype_counts = phenotype_counts,
  n_fold = 5,
  split = c(1, 0),
  min_total_obs = 40
)
```

## Loading ML matrices

```{r load-data}
# Load a generated matrix
ml_tibble <- loadMLInputTibble(
  parquet_path = "results/matrix/Cje_drug_CIP_genes_binary_sparse.parquet"
)

# Check dimensions
n_features <- getNumFeat(ml_tibble)
target_var <- getTargetVarName(ml_tibble)
```

## Main pipeline

The primary entry point is `runMLPipeline()`, which orchestrates the entire ML workflow:

```{r run-pipeline}
results <- runMLPipeline(
  ml_input_tibble = ml_tibble,
  model = "LR",
  split = c(0.6, 0.2),
  n_fold = 2,
  n_top_feats = 20,
  select_best_metric = "mcc",
  return_fit = TRUE,
  return_pred = TRUE,
  verbose = TRUE
)

# View results
results$performance_tibble
results$top_feat_tibble
```

## Step-by-step model building

For more control, use the individual tidymodels-based functions:

```{r split-data}
# Split data
data_split <- splitMLInputTibble(ml_tibble, split = c(0.6, 0.2), seed = 123)

library(rsample)
train_data <- training(data_split)
test_data <- testing(data_split)
```

```{r build-model}
# Build recipe (preprocessing)
recipe <- buildRecipe(train_data, use_pca = FALSE, pca_threshold = 0.95)

# Build logistic regression model
lr_model <- buildLRModel(multi_class = FALSE)

# Build workflow
wflow <- buildWflow(lr_model, recipe)

# Build tuning grid
grid <- buildTuningGrid(
  model = "LR",
  penalty_vec = 10^seq(-4, -1, length.out = 10),
  mix_vec = 0:5 / 5
)
```

```{r tune-fit}
# Tune hyperparameters
tune_res <- tuneGrid(wflow, data_split, grid, n_fold = 5)

# Select and fit best model
best_wflow <- selectBestModel(tune_res, wflow, select_best_metric = "mcc")
fit <- fitBestModel(best_wflow, train_data)

# Get fitted hyperparameters
fit_hps <- getFitHps(fit)

# Make predictions
predictions <- predict(fit, test_data)
```

## Performance metrics

```{r metrics}
# Individual metrics
nmcc <- calculatenMCC(predictions) # Normalized MCC (0-1 scale)
f1 <- calculateF1(predictions) # F1 score
bal_acc <- calculateBalAcc(predictions) # Balanced accuracy
auprc <- calculateAUPRC(predictions) # Area under PR curve
log2_apop <- calculateLog2APOP(predictions) # log2(AUPRC/prior)

# All metrics at once
all_metrics <- calculateEvalMets(predictions)

# Confusion matrix
conf_mat <- getConfusionMatrix(predictions)
```

| Metric | Function | Description |
|--------|----------|-------------|
| nMCC | `calculatenMCC()` | Normalized Matthews correlation coefficient (0-1) |
| F1 | `calculateF1()` | Harmonic mean of precision and recall |
| Balanced accuracy | `calculateBalAcc()` | Average of sensitivity and specificity |
| AUPRC | `calculateAUPRC()` | Area under precision-recall curve |
| log2(AUPRC/prior) | `calculateLog2APOP()` | Class-imbalance adjusted AUPRC |

## Feature importance

```{r feature-importance}
# Extract top features from fitted model by number
top_features <- extractTopFeats(fit, n_top_feats = 20)

# Or by proportion of variable importance
top_features <- extractTopFeats(fit, prop_vi_top_feats = c(0, 0.2))

# Columns: Variable, Importance, Sign
# Positive Sign = associated with resistance
# Negative Sign = associated with susceptibility
```

## Iterative feature elimination (IFE)

IFE iteratively removes top features to identify the most predictive subset:

```{r ife}
ife_results <- runIFE(
  ml_tibble,
  by_num = TRUE, # Remove by number of features
  by_vi = FALSE, # Or remove by variable importance
  percent_removal_vec = 10 * 1:9, # Remove 10%, 20%, ..., 90%
  mix_vec = 0, # L2 regularization for IFE
  return_feats = TRUE,
  verbose = TRUE
)

# Results include nMCC at each iteration
ife_results$ife_performance_tibble
ife_results$feats_removed

# Remove specific features manually
ml_tibble_reduced <- removeTopFeats(ml_tibble, top_features)
```

## Visualization

### Precision-recall curve

```{r plot-prc}
plotPRC(results$pred)
```

### Variable importance plot

```{r plot-vi}
plotTopFeatsVI(results$fit, n_top_feats = 10)
```

### Default evaluation plot

```{r plot-default-eval}
# Plot performance metric vs training proportion or CV folds
plotDefaultEval(
  default_eval_tibble,
  x_default_eval = "train_prop",
  y_default_eval = "avg_f1_score",
  xlab = "Training data proportion",
  ylab = "Average F1 score"
)
```

### Baseline comparison barplot

```{r plot-baseline}
# Run with shuffled labels for baseline
baseline_results <- runMLPipeline(
  ml_input_tibble = ml_tibble,
  model = "LR",
  split = c(0.6, 0.2),
  shuffle_labels = TRUE
)

# Compare performance
getBaselineComparisonBarplot(
  non_shuffled_label_results = results$performance_tibble,
  shuffled_label_results = baseline_results$performance_tibble
)
```

## Fisher's exact tests (baseline comparison)

As a non-ML baseline, run Fisher's exact tests with multiple testing correction:

```{r fisher}
# Complete Fisher pipeline
fisher_results <- runFishers(
  matrix_path = "results/matrix/Cje_drug_CIP_genes_binary_sparse.parquet",
  Q = 0.05,
  alternative = "two.sided",
  susceptible_label = "Susceptible",
  resistant_label = "Resistant"
)

# Or step-by-step:
encoded <- encodePhenotype(ml_tibble)
fisher_res <- runFisherTests(encoded$df, encoded$target, alternative = "two.sided")
fisher_res <- applyBenjaminiHochberg(fisher_res, Q = 0.05)
fisher_res <- computeFeatureFreq(encoded$df, fisher_res, encoded$target)
```
