---
title: "Introduction to amRml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to amRml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(amRml)
```

## Overview

`amRml` is part of a 3-package suite for predicting antimicrobial resistance (AMR) using machine learning models trained on bacterial genomic features. This vignette covers the complete API for data preparation, model training, evaluation, and visualization.

## Data preparation

### Loading data

The package expects Parquet files in long format with columns:

- `genome_id`: Unique identifier for each isolate
- `feature_id`: Feature name (e.g., gene, protein domain)
- `value`: Binary presence/absence (0 or 1)
- `genome_drug.resistant_phenotype`: "Resistant" or "Susceptible"

```{r load-data}
# Load Parquet file and convert to wide format
ml_data <- loadMLInputTibble("features.parquet")

# Get number of features
n_features <- getNumFeat(ml_data)

# Get target variable name
target_var <- getTargetVarName(ml_data)
```
### Splitting data

```{r split-data}
# Split data into train/validation/test sets
# split = c(0.6, 0.2) means 60% train, 20% validation, 20% test
data_split <- splitMLInputTibble(ml_data, split = c(0.6, 0.2), seed = 123)

# Access train and test sets
library(rsample)
train_data <- training(data_split)
test_data <- testing(data_split)
```

### Utility functions

```{r utils}
# Calculate minimum samples needed for valid ML
min_samples <- calculateMinSamples(
  n_fold = 5,
  split = c(0.6, 0.2),
  res_prop = 0.3
)

# Shuffle labels for baseline comparison
shuffled_data <- shuffleLabels(ml_data, seed = 123)
```

## Main pipeline

The primary entry point is `runMLPipeline()`, which orchestrates the entire ML workflow:

```{r run-pipeline}
results <- runMLPipeline(
  ml_input_tibble = ml_data,
  model = "LR", # "LR", "RF", or "BT"
  split = c(0.6, 0.2), # Train/validation proportions
  n_fold = 2, # CV folds (when validation = 0)
  prop_vi_top_feats = c(0, 1), # Feature importance range
  n_top_feats = NA, # Or specify exact number
  use_pca = FALSE, # Use PCA dimensionality reduction
  penalty_vec = 10^seq(-4, -1, length.out = 10), # LR penalty values
  mix_vec = 0:5 / 5, # LR mixture values (0=L2, 1=L1)
  min_n_vec = c(2, 6, 12), # RF/BT min_n values
  tree_vec = c(100, 500, 1000), # RF/BT tree counts
  select_best_metric = "mcc", # "mcc", "f_meas", "pr_auc", "bal_accuracy"
  seed = 123,
  shuffle_labels = FALSE, # For baseline comparisons
  return_tune_res = FALSE,
  return_fit = FALSE,
  return_pred = FALSE,
  verbose = TRUE
)

# Results
results$performance_tibble
results$top_feat_tibble
```

### Output structure

`runMLPipeline()` returns a list with:

- `performance_tibble`: Metrics including nMCC, F1, balanced accuracy, AUPRC
- `top_feat_tibble`: Ranked features with Variable, Importance, Sign
- `tune_res`: (optional) Tuning results
- `fit`: (optional) Fitted model object
- `pred`: (optional) Predictions with probabilities

## Step-by-step model building

For more control, use the individual tidymodels-based functions:

### Building models

```{r build-models}
# Build recipe (preprocessing)
recipe <- buildRecipe(train_data, use_pca = FALSE, pca_threshold = 0.95)

# Build model specification
lr_model <- buildLRModel(multi_class = FALSE)
rf_model <- buildRFModel()
bt_model <- buildBTModel()

# Build workflow
wflow <- buildWflow(lr_model, recipe)
```

### Tuning and fitting

```{r tune-fit}
# Build tuning grid
grid <- buildTuningGrid(
  model = "LR",
  penalty_vec = 10^seq(-4, -1, length.out = 10),
  mix_vec = 0:5 / 5
)

# For random forest or boosted tree:
# grid <- buildTuningGrid(
#   model = "RF",
#   n_feat = getNumFeat(train_data),
#   min_n_vec = c(2, 6, 12),
#   tree_vec = c(100, 500, 1000)
# )

# Tune hyperparameters
tune_res <- tuneGrid(wflow, data_split, grid, n_fold = 5)

# Select and fit best model
best_wflow <- selectBestModel(tune_res, wflow, select_best_metric = "mcc")
fit <- fitBestModel(best_wflow, train_data)

# Get fitted hyperparameters
fit_hps <- getFitHps(fit)
```

### Making predictions

```{r predict}
# Make predictions on test data
predictions <- predict(fit, test_data)

# Get confusion matrix
conf_mat <- getConfusionMatrix(predictions)
```

## Performance metrics

```{r metrics}
# Individual metrics
nmcc <- calculatenMCC(predictions) # Normalized MCC (0-1 scale)
f1 <- calculateF1(predictions) # F1 score
bal_acc <- calculateBalAcc(predictions) # Balanced accuracy
auprc <- calculateAUPRC(predictions) # Area under PR curve
log2_apop <- calculateLog2APOP(predictions) # log2(AUPRC/prior)

# All metrics at once
all_metrics <- calculateEvalMets(predictions)
```

| Metric | Function | Description |
|--------|----------|-------------|
| nMCC | `calculatenMCC()` | Normalized Matthews correlation coefficient (0-1) |
| F1 | `calculateF1()` | Harmonic mean of precision and recall |
| Balanced accuracy | `calculateBalAcc()` | Average of sensitivity and specificity |
| AUPRC | `calculateAUPRC()` | Area under precision-recall curve |
| log2(AUPRC/prior) | `calculateLog2APOP()` | Class-imbalance adjusted AUPRC |

## Feature importance

```{r feature-importance}
# Extract top features from fitted model
top_features <- extractTopFeats(
  fit,
  n_top_feats = 20 # Exact number
)

# Or by proportion of variable importance
top_features <- extractTopFeats(
  fit,
  prop_vi_top_feats = c(0, 0.2) # Top 20% of variable importance
)

# Feature importance columns: Variable, Importance, Sign
# Positive Sign = associated with resistance
# Negative Sign = associated with susceptibility
```

## Iterative feature elimination (IFE)

IFE iteratively removes top features to identify the most predictive subset:

```{r ife}
# Run IFE by removing percentage of features
ife_results <- runIFE(
  ml_data,
  by_num = TRUE, # Remove by number of features
  by_vi = FALSE, # Or remove by variable importance
  percent_removal_vec = 10 * 1:9, # Remove 10%, 20%, ..., 90%
  mix_vec = 0, # L2 regularization for IFE
  return_feats = TRUE, # Return features removed at each step
  verbose = TRUE
)

# Results include nMCC at each iteration
ife_results$ife_performance_tibble
ife_results$feats_removed # If return_feats = TRUE

# Remove specific features manually
ml_data_reduced <- removeTopFeats(ml_data, top_features)
```

## Visualization

### Precision-recall curve

```{r plot-prc}
# Plot precision-recall curve
prc_plot <- plotPRC(predictions)
print(prc_plot)
```

### Variable importance plot

```{r plot-vi}
# Plot top features by variable importance
vi_plot <- plotTopFeatsVI(fit, n_top_feats = 10)
print(vi_plot)
```

### Default evaluation plot

```{r plot-default-eval}
# Plot performance metric vs training proportion or CV folds
eval_plot <- plotDefaultEval(
  default_eval_tibble,
  x_default_eval = "train_prop", # or "n_fold"
  y_default_eval = "avg_f1_score", # or other metrics
  xlab = "Training data proportion",
  ylab = "Average F1 score"
)
print(eval_plot)
```

### Baseline comparison barplot

```{r plot-baseline}
# Compare model performance with and without shuffled labels
baseline_plot <- getBaselineComparisonBarplot(
  non_shuffled_label_results = results$performance_tibble,
  shuffled_label_results = baseline_results$performance_tibble
)
```

## Fisher's exact tests (baseline comparison)

As a non-ML baseline, run Fisher's exact tests with multiple testing correction:

```{r fisher}
# Complete Fisher pipeline
fisher_results <- runFishers(
  matrix_path = "features.parquet",
  Q = 0.05, # FDR threshold
  alternative = "two.sided",
  susceptible_label = "Susceptible",
  resistant_label = "Resistant"
)

# Or step-by-step:
ml_data <- loadMLInputTibble("features.parquet")
encoded <- encodePhenotype(ml_data)
fisher_res <- runFisherTests(encoded$df, encoded$target, alternative = "two.sided")
fisher_res <- applyBenjaminiHochberg(fisher_res, Q = 0.05)
fisher_res <- computeFeatureFreq(encoded$df, fisher_res, encoded$target)
```

## Complete workflow example

```{r workflow}
library(amRml)

# 1. Load and prepare data
ml_data <- loadMLInputTibble("species_drug_features.parquet")
message("Loaded ", nrow(ml_data), " samples with ", getNumFeat(ml_data), " features")

# 2. Run ML pipeline with logistic regression
results <- runMLPipeline(
  ml_input_tibble = ml_data,
  model = "LR",
  split = c(0.6, 0.2),
  n_top_feats = 50,
  select_best_metric = "mcc",
  return_fit = TRUE,
  return_pred = TRUE
)

# 3. View results
print(results$performance_tibble)
print(head(results$top_feat_tibble, 20))

# 4. Baseline comparison with shuffled labels
baseline_results <- runMLPipeline(
  ml_input_tibble = ml_data,
  model = "LR",
  split = c(0.6, 0.2),
  shuffle_labels = TRUE
)

# 5. Compare real vs baseline performance
cat("Real nMCC:", results$performance_tibble$nmcc, "\n")
cat("Baseline nMCC:", baseline_results$performance_tibble$nmcc, "\n")

# 6. Run iterative feature elimination
ife_results <- runIFE(
  ml_data,
  by_num = TRUE,
  percent_removal_vec = c(10, 25, 50, 75, 90),
  return_feats = TRUE
)

# 7. Generate plots
vi_plot <- plotTopFeatsVI(results$fit, n_top_feats = 15)
print(vi_plot)
```

## Model types

| Model | Function | Description | Hyperparameters |
|-------|----------|-------------|-----------------|
| Logistic regression | `buildLRModel()` | Interpretable, supports L1/L2/elastic net | `penalty`, `mixture` |
| Random forest | `buildRFModel()` | Ensemble method, robust | `trees`, `mtry`, `min_n` |
| Boosted tree | `buildBTModel()` | Gradient boosting | `trees`, `mtry`, `min_n` |

## Session info

```{r session-info, eval = TRUE}
sessionInfo()
```
