---
title: "Introduction to amRml"
output:
  BiocStyle::html_document:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Introduction to amRml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(amRml)
```

## Overview

`amRml` predicts antimicrobial resistance (AMR) from bacterial genomic features. It consumes a DuckDB produced by `amRdata` and produces ML matrices, tuned logistic regression models, per-genome predictions, feature importances, and Fisher's exact tests as a non-ML baseline.

This vignette uses the *Shigella flexneri* (`Sfl`) DuckDB bundled in `inst/extdata`.

```{r fixture}
fixture <- system.file("extdata", "Sfl_parquet.duckdb", package = "amRml")
out_dir <- file.path(tempdir(), "amRml_vignette")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

## Generating ML input matrices

`generateMLInputs()` reads the bug-level DuckDB (metadata + feature parquets) and writes one long-format sparse parquet per drug × feature × encoding combination into `out_path/matrix/`. With `stratify_by`, it additionally writes year- or country-stratified matrices into `matrix_year/` or `matrix_country/`.

```{r generate-matrices}
generateMLInputs(
  parquet_duckdb_path = fixture,
  out_path            = out_dir,
  n_fold              = 5,
  split               = c(1, 0),
  min_n               = 25,
  verbosity           = "minimal"
)

list.files(file.path(out_dir, "matrix"))[1:5]
```

For a classical train/validation/test split instead of cross-validation, set `n_fold = NULL` and pass a length-2 `split` of `c(train_prop, val_prop)`; the test proportion is the remainder.

```{r generate-splits, eval = FALSE}
generateMLInputs(
  parquet_duckdb_path = fixture,
  out_path            = out_dir,
  n_fold              = NULL,
  split               = c(0.7, 0.15),
  verbosity           = "debug"
)
```

## Loading a single matrix

The generated parquets are long-format sparse tibbles with these columns:

| Column | Description |
|--------|-------------|
| `genome_id` | Unique identifier for each isolate |
| `feature_id` | Feature name (gene, protein, domain, or struct) |
| `value` | Binary presence/absence (0/1) or count |
| `genome_drug.resistant_phenotype` | `"Resistant"` or `"Susceptible"` |

`loadMLInputTibble()` converts one of them to wide format (one row per genome, one column per feature) ready for ML.

```{r load-data}
matrix_path <- file.path(
  out_dir, "matrix", "Sfl_drug_AMP_genes_binary_sparse.parquet"
)

ml_tibble <- loadMLInputTibble(matrix_path)
n_features <- getNumFeat(ml_tibble)
target_var <- .getTargetVarName(ml_tibble)

c(n_features = n_features, target_var = target_var)
```

## Per-matrix pipeline

`runMLPipeline()` runs the train/tune/fit/predict pipeline on a single matrix in memory. Use it to iterate on one drug-feature combo before scaling to all of them.

```{r run-pipeline}
results <- runMLPipeline(
  ml_input_tibble    = ml_tibble,
  model              = "LR",
  split              = c(1, 0),
  n_fold             = 2,
  n_top_feats        = 20,
  penalty_vec        = 10^c(-3, -1),
  mix_vec            = c(0, 0.5, 1),
  select_best_metric = "mcc",
  return_fit         = TRUE,
  return_pred        = TRUE,
  verbose            = FALSE
)

results$performance_tibble
head(results$top_feat_tibble)
```

### Output structure

`runMLPipeline()` returns a named list:

**`performance_tibble`** — one row of model performance metrics:

| Column | Description |
|--------|-------------|
| `num_obs` | Number of observations |
| `res_prop` | Proportion of resistant samples |
| `n_feat` | Number of features |
| `model` | Model type (`"LR"`) |
| `train_prop`, `val_prop` | Train/validation split proportions |
| `fit_penalty`, `fit_mixture` | Fitted hyperparameters |
| `nmcc`, `f1`, `bal_acc`, `log2_apop` | Performance metrics |
| `run_time_sec` | Runtime in seconds |

**`top_feat_tibble`** — ranked feature importance:

| Column | Description |
|--------|-------------|
| `Variable` | Feature name |
| `Importance` | Variable importance score |
| `Sign` | Direction of effect (`POS` = associated with resistance, `NEG` = with susceptibility) |

**Optional outputs** (when `return_*` = TRUE):

- `tune_res` — tuning results from grid search
- `fit` — the fitted workflow object
- `pred` — predictions with `.pred_class`, `.pred_Resistant`, `.pred_Susceptible`

## Step-by-step model building

The builders below are what `runMLPipeline()` chains internally. Call them directly when you need control over any individual step.

```{r split-data}
data_split <- splitMLInputTibble(ml_tibble, split = c(0.6, 0.2), seed = 123)
train_data <- rsample::training(data_split)
test_data <- rsample::testing(data_split)
```

```{r build-model}
recipe <- buildRecipe(train_data, use_pca = FALSE)
lr_mod <- buildLRModel(multi_class = FALSE)
wflow <- buildWflow(lr_mod, recipe)

grid <- buildTuningGrid(
  model       = "LR",
  penalty_vec = 10^c(-3, -1),
  mix_vec     = c(0, 0.5, 1)
)
```

```{r tune-fit}
tune_res <- tuneGrid(wflow, data_split, grid, n_fold = 2)
best_wflow <- selectBestModel(tune_res, wflow, select_best_metric = "mcc")
fit <- fitBestModel(best_wflow, train_data)

preds <- predictML(fit, test_data)
```

## Performance metrics

`calculateEvalMets()` returns all of nMCC, F1, balanced accuracy, AUPRC, log2(AUPRC/prior), sensitivity, and specificity from a tibble of predictions + truth.

```{r metrics}
calculateEvalMets(preds)
getConfusionMatrix(preds)
```

## Feature importance

`extractTopFeats()` ranks features by absolute coefficient (for LR). Use `n_top_feats` for a fixed count or `prop_vi_top_feats` for a percentile range.

```{r feature-importance}
top_features <- extractTopFeats(fit, n_top_feats = 20)
head(top_features)
```

## Visualization

```{r plot-prc, fig.width = 5, fig.height = 4}
plotPRC(results$pred)
```

```{r plot-vi, fig.width = 5, fig.height = 4}
plotTopFeatsVI(results$fit, n_top_feats = 10)
```

For a baseline comparison against random labels, fit a shuffled-label pipeline and compare:

```{r plot-baseline, eval = FALSE}
shuffled <- runMLPipeline(
  ml_input_tibble = ml_tibble,
  model           = "LR",
  split           = c(1, 0),
  n_fold          = 2,
  shuffle_labels  = TRUE,
  return_pred     = TRUE
)

plotBaselineComparison(
  non_shuffled_label_results = results$performance_tibble,
  shuffled_label_results     = shuffled$performance_tibble
)
```

## Iterative feature elimination

`runIFE()` retrains the model after iteratively removing top-ranked features, helping identify the minimal predictive subset. It runs the pipeline once per percentile in `percent_removal_vec`

```{r ife, eval = FALSE}
ife_results <- runIFE(
  ml_tibble,
  by_num              = TRUE,
  by_vi               = FALSE,
  percent_removal_vec = 10 * 1:9,
  mix_vec             = 0,
  return_feats        = TRUE,
  verbose             = FALSE
)

ife_results$ife_performance_tibble
ife_results$feats_removed
```

`removeTopFeats()` strips a given set of features from a matrix tibble if you want to do this manually:

```{r remove-feats}
trimmed <- removeTopFeats(ml_tibble, head(top_features, 5))
ncol(ml_tibble) - ncol(trimmed)
```

## Fisher's exact tests (non-ML baseline)

`runFishers()` runs a Fisher's exact test of feature presence vs. phenotype for each feature, applies Benjamini–Hochberg correction, and computes per-class frequencies.

```{r fisher}
fisher_results <- runFishers(
  matrix_path       = matrix_path,
  Q                 = 0.05,
  alternative       = "two.sided",
  susceptible_label = "Susceptible",
  resistant_label   = "Resistant"
)

head(fisher_results)
```

```{r plot-fishers, fig.width = 5, fig.height = 4}
plotFishers(fisher_results, alpha = 0.05, label_top_n = 5)
```

## Training all models with `runMLmodels`

`runMLmodels()` trains a model on every matrix produced by `generateMLInputs()` and writes performance TSVs into `out_path/ML_performance/`, predictions into `ML_pred/`, and top features into `ML_top_features/`. Takes over an hour on Sfl with default settings.

```{r run-all-models, eval = FALSE}
runMLmodels(
  path            = out_dir,
  stratify_by     = NULL,
  LOO             = FALSE,
  cross_test      = FALSE,
  threads         = max(1L, parallel::detectCores() - 1L),
  split           = c(1, 0),
  n_fold          = 5,
  verbose         = TRUE,
  return_pred     = TRUE,
  use_saved_split = TRUE
)
```

## End-to-end: `runModelingPipeline`

For the full pipeline from a DuckDB to all outputs in one call:

```{r run-modeling-pipeline, eval = FALSE}
runModelingPipeline(
  parquet_duckdb_path = fixture,
  threads             = max(1L, parallel::detectCores() - 1L),
  n_fold              = 5,
  split               = c(1, 0),
  min_n               = 25,
  prop_vi_top_feats   = c(0, 1),
  pca_threshold       = 0.99,
  verbose             = TRUE,
  use_saved_split     = TRUE
)
```

## Session info

```{r session-info}
sessionInfo()
```
